---
title: "How Increased Bike Share Infrastucture Influenced a Historic Decline in Toronto Bicycle Thefts in 2024"
author: 
  - Tara Chakkithara
  - Sean Chua
date: today
date-format: long
format: html
number-sections: true
bibliography: references.bib
warning: false
message: false
echo: false
---

```{r}
library(tidyverse)
library(leaflet)
library(sf)
library(ggthemes)
library(ggimage)
library(waffle)
total_data <- read_csv('../data/analysis_data/analysis_data.csv')
data <- total_data |>
  filter(LONG_WGS84 != 0 & LAT_WGS84 != 0)
toronto <- st_read('../data/analysis_data/toronto.geojson', quiet = TRUE)
```

# Introduction

In the past decade, cycling has become a prominent mode of transportation in Toronto. By 2019, around 75 percent of households in the city owned a bicycle, and more than 40% of people reported using their bike for non-recreational purposes such as commuting to school or the office [@cityoftoronto2019cycling]. Despite being the safest city in Canada due to its low crime density [@mazloum2024ottawa], bike theft is a significant issue. Between 2010 and 2024, residents of Toronto faced a total financial loss of \$33,704,466 CAD from bike theft.

Although 2024 has seen the lowest number of bike thefts since 2013, as shown in @fig-trends, the recovery rate for stolen bikes remains low at 0.7 percent, making bike theft prevention a priority for cyclists. This analysis aims to explore the reasoning behind this   historical decline in bike thefts.

```{r, fig-trends, fig.cap="Bike Theft in Toronto from 2010 to 2024", fig.align='center'}

total_data |>
  filter(OCC_YEAR > 2010) |>
  count(OCC_YEAR) |>
  ggplot(aes(x = OCC_YEAR, y = n)) +
    geom_point(color = "#6C757D", size = 2) +  
    geom_line(color = "#17a2b8", size = 1) +
    scale_x_continuous(
      breaks = seq(min(total_data$OCC_YEAR[total_data$OCC_YEAR > 2010]), 
                   max(total_data$OCC_YEAR), 
                   by = 1)) +
    labs(
      x = "Year", 
      y = "Number of Incidents",
    ) +
    theme_minimal(base_size = 12) +  
    theme(
      plot.caption = element_text(hjust = 1, size = 10, color = "#777777"),
      axis.title = element_text(size = 12, color = "#333333"),  
      axis.text = element_text(size = 7, color = "#555555"),
      axis.title.x = element_text(size = 10, margin = margin(t = 25)),
      axis.title.y = element_text(size = 10, margin = margin(r = 25)),
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(color = "#dddddd", size = 0.5),
      plot.margin = margin(15, 15, 15, 15)
    )


```

# Characteristics of Stolen Bikes

Based on data from [Toronto's Public Safety Data Portal](https://data.torontopolice.on.ca/datasets/TorontoPS::bicycle-thefts-open-data/about), we determined that Mountain and Regular Bikes are stolen most often. Stolen bikes tend to have an average speed of 21 km/h and a typical market value of around \$1,000. Black bikes make up a significant majority of thefts, likely due to their inconspicuous nature, especially at night, as well as their widespread availability in the market. The majority of all bike thefts are tied to other primary offenses, where the bike is either used as a getaway vehicle or stolen opportunistically. Around 44% of thefts are standalone incidents where bike theft is the primary crime as seen in @fig-primary.

```{r}
#| label: fig-primary
#| fig-cap: "Categories of Bike Theft"
#| fig-align: center

primary <- c("THEFT OVER - BICYCLE", "THEFT OF EBIKE UNDER $5000", "THEFT UNDER - BICYCLE",
             "THEFT OF EBIKE OVER $5000")
total_data |>
  mutate(
    TYPE = case_when(
      PRIMARY_OFFENCE %in% primary ~ "Primary Offense",
      TRUE ~ "Non Primary Offense"
    )
  ) |>
  ggplot(aes(x = "", fill = TYPE)) + 
  geom_bar(width = 1, stat = "count") +
  coord_polar(theta = "y") +
  scale_fill_manual(values = c("Primary Offense" = "#003366", "Non Primary Offense" = "#FFCC00")) +
  theme_void() +
  theme(
    legend.title = element_blank()
  )
```

# Spatial Distributions of Bike Theft 

As shown in @fig-neighbourhood, the Yonge-Bay Corridor has the highest number of bike thefts in Toronto, with 2,141 recorded thefts since 2010. In contrast, the Maple Leaf area has experienced the fewest thefts, with only 8 incidents over the same period. The Yonge-Bay Corridor is characterized by a population in which 51% are adults aged 25 to 54. The largest income group in the area earns less than 10,000 CAD, accounting for 30% of the population aged 15 and over. Additionally, 98% of residents live in apartments [@yongebay2016]. In comparison, Maple Leaf has a population approximately half the size of Yonge-Bay. 46% of its residents live in single-detached homes, and 40% of the working population earns between 20,000 and 50,000 CAD [@mapleleaf2016].

According to @fig-apt, apartments are more vulnerable to bike thefts compared to detached homes. Therefore, it is not surprising that the Yonge-Bay Corridor experiences higher rates of bike thefts. Additionally, the potential of selling stolen bicycle parts may provide a significant financial incentive to some residents, given the high levels of poverty in the region.


Bike thefts occur more often near apartments since apartments often have communal bike storage areas which may not be as secure as individual garages in detached homes. In fact, in a 2018 article, it was written that “for the first time since Toronto police began collecting data, the number of bikes stolen from apartments, condos and their private storage rooms has surpassed those stolen on the street [@aptthefts].” Such areas are also easily accessed by multiple residents and increase the chances of theft occurring.

```{r}
data <- data |>
  select(HOOD_158, LONG_WGS84, LAT_WGS84) |>
  group_by(HOOD_158) |>
  summarise(num_thefts = n()) |>
  filter(!is.na(HOOD_158)) |>
  rename(AREA_LONG_CODE = HOOD_158)
```

```{r}
merged_data <- toronto |>
  left_join(data, by = "AREA_LONG_CODE")
```

```{r}
#| label: fig-neighbourhood
#| fig-cap: "The Distribution of Bicycle Thefts Across Toronto Neighbourhoods"
#| fig-align: center

ggplot(data = merged_data) +
  geom_sf(aes(fill = num_thefts), color = NA, size = 0.3) +
  scale_fill_viridis_c(
    option = "C",
    trans = "log",
    direction = -1,
    breaks = c(20, 50, 150, 400, 1000),
    guide = guide_colorbar(
      barheight = unit(0.2, units = "cm"),
      barwidth = unit(5, units = "cm"),
      label.position = "bottom",
      title.position = "top",
      nrow = 1,
      label.theme = element_text(size = 8, color = "#4e4d47"),
      title.theme = element_text(
        size = 10,
        color = "#4e4d47",
        face = "plain"
      ),
      ticks = FALSE
    )
  ) +
  theme_minimal() +
  theme(
    legend.position = "right",
    legend.title = element_text(face = "bold", size = 12),
    legend.text = element_text(size = 10),
    legend.key.size = unit(0.7, "cm"),
    plot.title = element_text(
      face = "bold",
      size = 16,
      hjust = 0.5,
      color = "black"
    ),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    axis.title = element_blank(),
    panel.grid = element_blank(),
    plot.margin = margin(10, 10, 10, 10)
  ) +
  theme(legend.position = 'bottom') +
  labs(fill = 'Number of Bike Thefts')
```

From 2010 to 2024, most bike thefts have occurred in outdoor areas, such as parks, hiking trails, and streets. These locations are particularly vulnerable due to the lack of dedicated bike storage infrastructure. In contrast, transit areas and educational institutions experience significantly lower rates of bike theft, as shown in @fig-apt. These areas often offer secure bike racks or lockers, which provide cyclists with a place to lock their bikes. Additionally, these locations tend to attract higher foot traffic, which can deter potential thieves.

```{r}
waffle_frame <- total_data |>
  filter(OCC_YEAR > 2010) |>
  count(PREMISES_TYPE)
```

```{r}
waffle_data <- c("Apartment" = 8849, "Commercial" = 4278, "Education" = 1578, "House" = 5333, "Other" = 4473, "Transit" = 774, "Outside" = 10834)
```

```{r}
#| label: fig-apt
#| fig-cap: "The Distribution of Bicycle Thefts Across Location Types"
#| fig-align: center

total <- sum(waffle_data)
scaled_data <- round(waffle_data / total * 100)

plot <- waffle(scaled_data, rows = 10, 
               legend_pos = "right",
               colors = c("Apartment" = "#4B9CD3",  
                          "Commercial" = "#3F6F53", 
                          "Education" = "#D1A23A",  
                          "House" = "#9F7B87",       
                          "Other" = "#D64A4A",       
                          "Transit" = "#F06D4D",     
                          "Outside" = "#7E9A92"))    

plot + 
  theme_minimal() +  
  theme(
    legend.title = element_blank(),
    legend.text = element_text(size = 10),
    legend.position = "right",  
    axis.text = element_blank(),  
    axis.ticks = element_blank(), 
    plot.caption = element_text(size = 10, color = "gray50", hjust = 0.5) 
  ) + 
  labs(caption = "1 Tile Represents 391 Thefts")
```


# Timing of Thefts 

As seen in @fig-hour, the bike thefts mostly peak at around 7:00PM. The decreasing visibility at this time makes it easier for thieves to operate undetected. Additionally, commuters tend to return home from work or school during this time, leading to a higher volume of parked bikes on streets and other public areas.  


```{r}
#| label: fig-hour
#| fig-cap: "Density of Bike Thefts By Hour of Day"
#| fig-align: center

total_data |>
  ggplot(aes(x = OCC_HOUR)) +
  geom_density(
    fill = "#4C9F70",       
    color = "#2A5D3D",      
    size = 1,               
    adjust = 1.5,          
    alpha = 0.6             
  ) +
  labs(
    x = "Hour of Day", 
    y = "Density",  
  ) +
  theme_minimal(base_size = 14) + 
  theme(
    axis.title = element_text(size = 12, color = "#333333"),  
    axis.text = element_text(size = 10, color = "#555555"),
    axis.title.x = element_text(size = 10, margin = margin(t = 25)),
    axis.title.y = element_text(size = 10, margin = margin(r = 25)),
    axis.ticks = element_line(color = "gray"), 
    panel.grid.major = element_line(color = "lightgray", linetype = "dotted"),  
    panel.grid.minor = element_blank(), 
    plot.caption = element_text(size = 12, color = "gray", hjust = 1)
  )

```


# Decreased Theft in 2024

As shown in @fig-apt, outdoor areas are particularly vulnerable to bike theft, largely due to the absence of secure bike storage infrastructure. However, starting around 2020, this trend began to shift with the expansion of Bike Share Toronto, which correlates with the observed decline in bike thefts, as seen in @fig-trends.

The Bike Share Toronto program reached its highest ridership in May 2020 [@toronto2020], and this growth has been instrumental in expanding secure bike storage options throughout the city. The availability of these secure bike stations, managed by the Toronto Parking Authority, has played a key role in improving bike security.
People are now more likely to use bike share bikes and keep their own bikes at home for recreational use. 

By 2024, bike sharing has become one of the fastest-growing forms of public transportation in the city [@CBC2024]. One of the key reasons for the drop in bike thefts is that the bikes in the program are owned by the Toronto Parking Authority and are stored in locked, company-managed racks. This system significantly reduces the risk of theft compared to privately owned bikes, which are more vulnerable in unsecured outdoor areas. The rise of bike sharing and its infrastructure has thus contributed to a safer cycling environment across Toronto.


# Limitations

Limitations of our analysis arise from the scope and nature of the dataset used. Our analysis is based solely on data related to bike thefts, without incorporating external factors that could offer a more comprehensive understanding of broader crime trends.

For instance, after conducting additional research, we found that bike thefts have been one of the only crimes to show a decline in recent years. In contrast, car theft has increased significantly, with insurance companies covering over 1 billion CAD in claims for car thefts in 2024 alone, as reported by @yousif2024. This broader trend of rising vehicle thefts is not reflected in our dataset, which may result in an unrealistic view of crime trends.

Additionally, the declining number of police officers entering the force could impact both the accuracy and consistency of crime reporting. Since 1999, only 37 officers have joined the Toronto Police Service, @fanfair2024, even as Toronto’s population grew by over 100,000 people in 2023 @clayton2024. This disparity between population growth and police staffing may contribute to challenges in crime reporting and response, potentially affecting the reliability of the data.

While our analysis provides valuable insights into bike theft patterns, it does not account for the broader context of rising vehicle thefts or the limitations posed by police staffing and reporting. A more comprehensive analysis would include additional data on other types of crime, law enforcement trends, and demographic changes to paint a more accurate picture of crime dynamics in Toronto.

# Conclusion

Toronto currently faces significant challenges related to income and housing, which often act as a motivators for crime such as bike thefts. However, the city's efforts to provide affordable bike sharing has had a positive impact. Through these initiatives, Toronto is proactively improving the quality of life, and at the same time, it is reducing crime as a result of increased infrastructure.

# References
